Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-21 paper/paper.tex
Revise section 3

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }

The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance.

\Rref{I-Pi} is the most interesting. The precondition consists of a
mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
subject to the constraint that $[[forall a, (# a in S implies Interp I
i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
the derivation, the statements from the precondition are equivalent to
the conjunction of the following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
\end{itemize}
In general, let $R$ be a binary relation over the sets $[[A]]$ and
$[[B]]$, it easy to verify the following equivalence, which is used to
justify the skolemization process~\citep{skolemization}.
\scw{Is this true constructively? If not, is that ok?}
\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
  \in R\]
The left-hand side is equivalent to the right-hand side, but has no
mentioning of any functions.
In the context of \rref{I-Pi}, through
the equivalence, we can more easily tell what's happening: the
function type $[[Pi A B]]$ has a semantic interpretation if its input
type $[[A]]$ can be semantically
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has a semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the function $[[F]]$, we can use $[[F a]]$ to retrive the
witness in the precondition for $[[a in S]]$.
With the alternative representation that is free of the function $[[F]]$, we
might want to reformulate our logical relation in the following way.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[F]]$ each time we apply it. Therefore, in
Lemma~\ref{lemma:piintroalt}, we show the admissibility of
\rref{I-PiAlt} a posteriori and use that as a more convenient
introduction rule for the rest of our proofs.

@@

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi A0
B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    These properties are exactly the ones from the conclusion.
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
$[[Interp I i B S]]$ also holds. The following property shows that
preservation holds in the usual forward direction too.
\begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.
  \begin{description}
  \item[\Rref{I-Void}:] There exists some $[[B]]$ such that $[[Void =>
    B]]$. By inverting the derivation of $[[Void => B]]$, $[[B]]$ must
    be $[[Void]]$ and the result trivially follows.
  \item[\Rref{I-Bool, I-Set}:] Similar to the case for \rref{I-Void}.
  \item[\Rref{I-Eq}:] We know that $[[Interp I i a ~ b : A { p | p =>+
      refl , a <=> b  }]]$ and, by inverting the derivation of
    parallel reduction, $[[a => a0]]$, $[[b => b0]]$, $[[A => A0]]$
    for some $[[a0]]$, $[[b0]]$, and $[[A0]]$. Our goal is to show
    that $[[Interp I i a0 ~ b0 : A {p | p =>+ refl, a <=> b}]]$. By
    \rref{I-Eq}, we already know that $[[Interp I i a0 ~ b0 : A {p | p
      =>+ refl, a0 <=> b0}]]$ and therefore it suffices to show that
    the sets $[[{p | p =>+ refl, a <=> b}]]$ and $[[{p | p =>+ refl,
      a0 <=> b0}]]$ are equal. Equivalently, it suffices to show that
    $[[a <=> b]]$ if and only if $[[a0 <=> b0]]$. By definition, from
    $[[a => a0]]$ and $[[b => b0]]$, we derive $[[a <=> a0]]$ and $[[b
    <=> b0]]$. The result then immediately follows from the fact that
    coherence is an equivalence relation.
  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
    such that:
    \begin{itemize}
    \item $[[Interp I i Pi A B { b | forall a, (# a in S implies b a in F
        a #) }]]$
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    There exists some $[[A0]]$ and $[[B0]]$ such that $[[A => A0]]$ and
    $[[B => B0]]$. Our goal is to show that $[[Interp I i Pi A0 B0 { b | forall a, (# a in S implies b a in F
      a #) }]]$. By \rref{I-Pi}, it suffices to show:
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \end{itemize}
    Since $[[A => A0]]$ and $[[B {a} => B0 {a}]]$ forall $[[a]]$, the
    latter of which follows from Lemma~\ref{lemma:parsubst}, the above
    conditions follow immediately from the induction hypothesis.
  \item[\Rref{I-Red}:] There exists $[[B0]]$ and $[[S]]$ such that $[[A => B0]]$ and
    $[[Interp I i B0 S]]$. Given an arbitrary $[[B]]$ such that $[[A
    => B]]$, our goal is to show that $[[Interp I i B S]]$. By the diamond
    property of parallel
    reduction (Lemma~\ref{lemma:pardiamond}), there exists some term
    $[[C]]$ such that $[[B0 => C]]$ and $[[B => C]]$. By the induction
    hypothesis, we derive $[[Interp I i C S]]$ from $[[Interp I i
    B0 S]]$. By \rref{I-Red} and
    $[[B => C]]$, we conclude that $[[Interp I i B S]]$.
  \end{description}
From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
derive the following corollary that two coherent types have the same
interpretation.
\scw{I'm not used to this lemma being called irrelevance. Intuitively,
it says that the definition of the logical relation respects
definitional equality. }
\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
  \label{lemma:logrelcoherence}
  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
\end{corollary}
\scw{Need a transition here}
Next, we show that $[[Interp I i A S]]$ is in fact a partial function.
\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
  \label{lemma:logreldeter}
  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
\end{lemma}
\begin{proof}
  We start by inducting over the derivation of the first premise $[[Interp I i A
  S0]]$.
  \begin{description}
  \item[\Rref{I-Void}:] We know that $[[Interp I i Void
    emptyset]]$. Given $[[Interp I i Void S1]]$, our goal is to show that
    $[[emptyset = S1]]$. This is immediate by applying the $[[Void]]$ case of
    Lemma~\ref{lemma:interpinv} to $[[Interp I i Void S1]]$.
  \item[\Rref{I-Bool, I-Eq, I-Set}:] Similar to the \rref{I-Void} case
    by applying the matching case of Lemma~\ref{lemma:interpinv}
    to $[[Interp I i A S1]]$.
  \item[\Rref{I-Pi}:] There exists $[[S]]$ and $[[F]]$
    such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    Our goal is to show that given $[[Interp I i Pi A B S1]]$, we have
    $[[S1 = { b | forall a, (# a in S implies b a in F
      a #) }]]$.
    By the function case of Lemma~\ref{lemma:interpinv}, there exists
    some $[[S0]]$ and $[[F0]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S0 ]]$
    \item $[[F0 in S0 -> PowerSet STm]]$
    \item
      $[[forall a, (# a in S0 implies Interp I i B { a } F0 a #)]]$
    \item
      $[[S1 = { b | forall a, (# a in S0 implies b a in F0 a #) }]]$
    \end{itemize}
    It suffices to show that $[[S = S0]]$ and $[[F = F0]]$. The
    equality $[[S = S0]]$ immediately follows from the induction
    hypothesis since $[[Interp I i A S]]$ and $[[Interp I i A
    S0]]$. Therefore, functions $[[F]]$ and $[[F0]]$ have the same
    domain and codomain and thus it suffices to show that forall $[[a in S]]$,
    $[[F a = F0 a]]$. Suppose $[[a in S]]$, we must have $[[Interp I
    i B {a} F a]]$ and $[[Interp I i B {a} F0 a]]$ from the two
    $\forall$-quantified statements above. The equality $[[F a = F0 a]]$ then
    immediately follows from the induction hypothesis.
  \item[\Rref{I-Red}:] There exists some $[[B]]$ such that $[[A =>
    B]]$ and $[[Interp I i B S0]]$. Our goal is to show that given
    $[[Interp I i A S1]]$, we have $[[S0 = S1]]$. By
    Lemma~\ref{lemma:interppreservation}, from $[[Interp I i A S1]]$
    and $[[A => B]]$, we have $[[Interp I i B S1]]$. From the
    induction hypothesis, we can conclude that $[[S0 = S1]]$ since
    $[[Interp I i B S0]]$ and $[[Interp I i B S1]]$.
  \end{description}

Lemma~\ref{lemma:logreldeter} enables us to show the admissibility of
\rref{I-PiAlt} and its related inversion principle.
\begin{lemma}[Pi Intro Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
  \label{lemma:piintroalt}
  Suppose the following two statements hold:
  \begin{itemize}
  \item $[[Interp I i A S]]$
  \item $[[forall a, a in S implies (# exists S0 , Interp I i B {a} S0 #)]]$
  \end{itemize}
  Then we have $[[Interp I i Pi A B { b | forall a, (# a in S , forall
    S0, (# Interp I i B {a} S0,  b a in F a #) #) }]]$
\end{lemma}
\begin{proof}
  Let $[[F]]$ be the relation defined as follows:
  \[ (a,S) \in [[F]] \iff [[Interp I i B {a} S]] \]
  By the second bullet from the premise and
  Lemma~\ref{lemma:logreldeter}, $[[F]]$ is a function that is total
  on the set $[[S]]$. The conclusion then trivially follows from
  \rref{I-Pi}.
\end{proof}

\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}

% The combination of Lemmas~\ref{lemma:piintroalt} and
% \ref{lemma:piinvalt} reveals what we truly want to capture with
% \rref{I-Pi}. A function type is semantically well-defined if its input
% type is well-defined, and its output type is defined for each valid
% inahbitant of its input type.

>>>>

The complexity of applying proof by logical relation to dependent types stem
from the fact that the logical relation is much harder to define. In
simply typed languages, the logical relation is defined as a recursive
function over the type $[[A]]$. In dependent types, the type
$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
this type, we need to first reduce it to $[[Bool]]$. However, we
cannot write a function that performs the reduction because we do not
know the termination of well-typed terms a priori. As a result, we
define the logical relation as an inductively defined relation,
reminiscent of how we specify the reduction graph of a partial
function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Par} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Par} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Par}
and \rref{I-Bool}.

\Rref{I-Pi} is the most interesting. To explain what it means, we
need to perform a few steps of transformations to its precondition.
The precondition consists of a
mysterious relation $[[R]]$ over the set $[[S]]$, an interpretation
of the type $[[A]]$, and returns a subset of terms. $[[R]]$ is further
subject to the following two constraints:
\begin{itemize}
\item $[[forall a, (# exists S0  , ( a , S0 ) in R #)]]$
\item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
\end{itemize}
The first condition allows us to induce a function $[[F in S ->
PowerSet STm]]$ such that $[[(a , F a) in R]]$. Then we can easily
show that $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$.

For the other direction, given a function $[[F]]$ satisfying the
property $[[forall a, (# a in S implies Interp I i B { a } F a #) ]]$, it
is possible to go in the other direction and define a relation $[[R]]=[[
{ ( a , F a )  | a in S }]]$. It is easy to verify that $[[R]]$ satisfy
the the constraint required in the precondition of \rref{I-Pi}.

Furthermore, the skolemization process allows us to conclude the
following equivalence:
\[ \exists F, [[forall a, (# a in S implies Interp I i B { a } F a #)]] \iff [[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]  \]
% Ignoring the content below the horizontal bar of
% the derivation, we claim the statements from the precondition are equivalent to
% the conjunction of the following statements:
% \begin{itemize}
% \item $[[Interp I i A S]]$
% \item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
%   [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
% \end{itemize}
% In general, let $R$ be a binary relation over the sets $[[A]]$ and
% $[[B]]$, it easy to verify the following equivalence, which is used to
% justify the skolemization process~\citep{skolemization}.
% \scw{Is this true constructively? If not, is that ok?}
% \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
%   \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
%   \in R\]
% The left-hand side is equivalent to the right-hand side, but has no
% mentioning of any functions.
The precondition of \rref{I-Pi} therefore can be formulated as the
following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $[[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]$
\end{itemize}
From this alternative formulation, we can more easily tell what is
happening in \rref{I-Pi}: the
function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[R in S * PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the relation $[[R]]$ in scope, we can refer to the
witness sets $[[S0]]$ such that $[[Interp I i B {a} S0]]$ for $[[a in S]]$.
With the alternative representation that is free of the relation $[[R]]$, we
might want to reformulate our logical relation in the following more
concise but equivalent form, whose admissibility follows directly from
\rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}

Unfortunately, \rref{I-PiAlt} not only violates the syntactic strict
positivity constraint required in proof assistants, but
is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt}.

@@

First, we prove a family of simple properties, which we refer to as
inversion principles for our logical relation. Given $[[Interp I i A
S]]$ where $[[A]]$ is in some head form such as $[[Bool]]$ or $[[Pi A0
B0]]$, the inversion lemma allows us to say something about the set
$[[S]]$. Its proof is simple, but we sketch out the case for
functions to help readers confirm their understanding of \rref{I-Pi}.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}} If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}} If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}} If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}} If $[[Interp I i Pi A B S1]]$, then there exists $[[S]],[[R]]$ such that:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[R in S * PowerSet STm]]$
    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
    \end{itemize}
  \item\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}} If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  As mentioned earlier, we only show the inversion property for the
  function type.
  We start by inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S1]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S1]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[R]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[R in S * PowerSet STm]]$
    \item $[[forall a, (# a in S implies (# exists S0 , (a , S0) in R #) #)]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B0 { a } S0 #)]]$
    \item $[[S1 = { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$
    \end{itemize}
    By inverting the derivation of $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[R]]$ additionally satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
    \end{itemize}
    These properties are exactly what we need to finish the proof.
  \end{description}
\end{proof}

\Rref{I-Red} bakes into the logical relation the backward preservation
property. That is, given $[[Interp I i A S]]$, if $[[B =>+ A]]$, then
$[[Interp I i B S]]$ also holds. The following property shows that
preservation holds in the usual forward direction too.
\begin{lemma}[Preservation of the logical relation\footnote{\dotv{semtyping.v}{InterpExt\_preservation}}]
  \label{lemma:interppreservation}
  If $[[Interp I i A S]]$ and $[[A => B]]$, then $[[Interp I i B S]]$.
\end{lemma}
\begin{proof}
  We carry out the proof by induction over the derivation of $[[Interp
  I i A S]]$.

  The only interesting case is \Rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B S]]$.

  The remaining cases all fall from induction hypotheses and basic
  properties about convertibility and parallel reduction we have
  established in Section~\ref{sec:spec}.
\end{proof}
From Lemma~\ref{lemma:interppreservation} and \rref{I-Red}, we can easily
derive the following corollary that two convertible types can always interpret
into the same set. We adopt the terminology from \citet{martin-lof-a-la-coq}
and refer to this property as irrelevance.
\begin{corollary}[Irrelevance of logical relation\footnote{\dotv{semtyping.v}{InterpUnivN\_Coherent}}]
  \label{lemma:logrelcoherence}
  If $[[Interp I i A S]]$ and $[[A <=> B]]$, then $[[Interp I i B S]]$.
\end{corollary}
% \scw{Need a transition here}

Since the definition of our logical relation is an inductive relation,
it is not immediately obvious why each type $[[A]]$ can only uniquely
corresponds to one set $[[S]]$. The following lemma shows that our
logical relation is indeed functional.
\begin{lemma}[Logical relation is functional\footnote{\dotv{semtyping.v}{InterpExt\_deterministic}}]
  \label{lemma:logreldeter}
  If $[[Interp I i A S0]]$ and $[[Interp I i A S1]]$, then $[[S0 = S1]]$.
\end{lemma}
\begin{proof}
  The proof proceeds by induction over the derivation of the first
  premise $[[Interp I i A S0]]$.
  All cases that are not \rref{I-Red} follow immediately from the
  Lemma~\ref{lemma:interpinv}.

  For \rref{I-Red}, we are given that there exists some $[[B]]$ such
  that $[[A => B]]$ and $[[Interp I i B S0]]$. Our goal is to show
  that given $[[Interp I i A S1]]$ for some $[[S1]]$, we have $[[S0 =
  S1]]$. By the preservation property
  (Lemma~\ref{lemma:interppreservation}),
  we know that $[[Interp I i B S1]]$ since $[[A => B]]$. The statement
  $[[S0 = S1]]$ then immediately follows from the induction hypothesis.
\end{proof}

Lemma~\ref{lemma:logreldeter} enables us to show the following
improved inversion lemma for function types whose statement is free of
the relation $[[R]]$, analogous to the admissible rule \rref{I-PiAlt}.
\begin{lemma}[Pi Inv Alt\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv\_nopf}}]
  \label{lemma:piinvalt}
  Suppose $[[Interp I i Pi A B S]]$, then there exists some $[[S0]]$
  such that the following constraints hold:
  \begin{itemize}
  \item $[[Interp I i A S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , Interp I i B {a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
    S1, (# Interp I i B {a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Immediate from Lemmas~\ref{lemma:interpinv} and \ref{lemma:logreldeter}.
\end{proof}
